---
title: "Getting and Cleaning Data Project"
author: "DA"
date: "July 22, 2015"
output: html_document
---
#**Introduction**
##Purpose
This document is the codebook describing the steps involved generating a tidy data set for Getting and Cleaning Data Project solution.
The resources are available on Github [here](https://github.com/dantohe/datasciencecoursera/tree/master/GettingandCleaningData/Project).

##Files

1. run_analysis.R - contains the script that will generate the data (intermediary data (point 4 of the assignment AND also the tidy data set required at point#5).

2. independent_tidy_data_set.txt - the tidy data set which is the solution for point 5 of the project requirement

3. CodeBokk - the code book for the tidy data set  - describes the features from independent_tidy_data_set.txt

4. PROJECT_README.txt - this file

##Running the script and generating the tidy data set

To run and generate the tidy data set follow the next step:
1. download the original data set from [here](https://d396qusza40orc.cloudfront.net/getdata%2Fprojectfiles%2FUCI%20HAR%20Dataset.zip)
2. Unpack the content of the zip file 
3. Go inside the unpacked folder and look for the data root: "UCI HAR Dataset"
4. Place there (in "UCI HAR Dataset" directory) run_analysis.R script - the script needs to be at the same level with "test" and "train" directories 
5. Run the script - it is recommended to run the script inside the R Studio - by setting the work directory to the directory where the run_analysis.R  was placed. To run just issue the following command inside the "R Studio" console:
source('run_analysis.R')
6. By running the above a script independent_tidy_data_set.txt will be created.


#**Data processing steps**

##Loading the raw data
This step just load all the files into variables (data frames)
```{r, echo=TRUE}
cat ("Loading raw data\n")
# 'features.txt': List of all features.
features_raw <- read.table(file = "features.txt")
cat(paste("features.txt - loaded ",nrow(features_raw)," rows\n"))
#'activity_labels.txt': Links the class labels with their activity name.
activity_labels_raw <- read.table(file = "activity_labels.txt")
cat(paste("activity_labels.txt - loaded ",nrow(activity_labels_raw)," rows\n"))
# 'subject_train.txt': List of test subjects.
training_subjects_raw <- read.table(file = "train/subject_train.txt")
cat(paste("train/subject_train.txt - loaded ",nrow(training_subjects_raw)," rows\n"))
# - 'train/X_train.txt': Training set.
training_set_raw  <- read.table(file = "train/X_train.txt")
cat(paste("train/X_train.txt - loaded ",nrow(training_set_raw)," rows\n"))
# - 'train/y_train.txt': Training labels.
training_set_labels_raw  <- read.table(file = "train/y_train.txt")
cat(paste("train/y_train.txt - loaded ",nrow(training_set_labels_raw)," rows\n"))
# 'subject_test.txt': List of test subjects.
test_subjects_raw <- read.table(file = "test/subject_test.txt")
cat(paste("test/subject_test.txt - loaded ",nrow(test_subjects_raw)," rows\n"))
# - 'test/X_test.txt': Test set.
test_set_raw  <- read.table(file = "test/X_test.txt")
cat(paste("test/X_test.txt - loaded ",nrow(test_set_raw)," rows\n"))
# - 'test/y_test.txt': Test labels.
test_set_labels_raw  <- read.table(file = "test/y_test.txt")
cat(paste("test/y_test.txt - loaded ",nrow(test_set_labels_raw)," rows\n\n"))
```

##Combine train and test data
This step creates data frames for test and train data by adding activity and subject columns to the waw data. 
At the end combines the train and test data to create a single large data frames containing all records (with the activity and subject columns added).
You can also embed plots, for example:

```{r, echo=TRUE}

#this function creates data frames for test and train data by adding activity and subject columns to the waw data 
#at the end combines the train and test data to create a single large data frames containing all records (with the activity and subject columns added)
cat("Preparing training data\n")
# PROCESSING TRAINING SET
#change the column names for the data frames
colnames(training_subjects_raw) <- c("subject")
colnames(training_set_labels_raw) <- c("activity")
#adding the features names as the column names
colnames(training_set_raw) <- features_raw$V2

#combine (cbind) the training data to have a data frame having the following format:
#SUBJECT, ACTIVITY, FEATURES (all the features)
saf_training_set_raw <- cbind(training_subjects_raw,training_set_labels_raw, training_set_raw)
cat(paste("Training data has ",ncol(saf_training_set_raw)," columns and ",nrow(saf_training_set_raw)," rows\n" ))

cat("Preparing test data\n")
# PROCESSING TRAINING SET
#change the column names for the data frames
colnames(test_subjects_raw) <- c("subject")
colnames(test_set_labels_raw) <- c("activity")
#adding the features names as the column names
colnames(test_set_raw) <- features_raw$V2

#combine (cbind) the test data to have the following format:
#SUBJECT, ACTIVITY, FEATURES (all the features)
saf_test_set_raw <- cbind(test_subjects_raw,test_set_labels_raw, test_set_raw)
cat(paste("Test data has ",ncol(saf_test_set_raw)," columns and ",nrow(saf_test_set_raw)," rows\n" ))

#combine (rbind) the test and training data - the result will be a 
combined_test_and_training_raw <- rbind(saf_training_set_raw, saf_test_set_raw)
cat(paste("Combined test and training data has ",ncol(combined_test_and_training_raw)," columns and ",nrow(combined_test_and_training_raw)," rows\n" ))
```

##Filter features
The raw data set has 561 features - we need to filter down to only those encoding mean and std deviation of some features. 
The following function does exactly that - the output will be a data frame having subject, activity and 48 columns containing only the desired features.
```{r, echo=TRUE}

cat("Filtering features - keep only subject, activity and 48 that are means and std.deviation\n")
#only some columns are of interest 
#collect the variable names
filtered_names <- grep(".+(mean\\(\\)|std\\(\\)).+", as.character(features_raw$V2), ignore.case = TRUE, perl=TRUE)
kept_variable_names<-as.character(features_raw$V2)[filtered_names]
kept_variable_names <- c("subject", "activity",  kept_variable_names )
#apply the filter 
filtered_variable_data_set <- combined_test_and_training_raw[kept_variable_names]
cat(paste("Data set filterd for features has ",ncol(filtered_variable_data_set)," columns and ",nrow(filtered_variable_data_set)," rows\n" ))
```

##Change activity names
Changes the activity names to replace the numbers with descriptive activity names.
```{r, echo=TRUE}
filtered_variable_data_set$activity[filtered_variable_data_set$activity=="1"]<-"WALKING"
filtered_variable_data_set$activity[filtered_variable_data_set$activity=="2"]<-"WALKING_UPSTAIRS"
filtered_variable_data_set$activity[filtered_variable_data_set$activity=="3"]<-"WALKING_DOWNSTAIRS"
filtered_variable_data_set$activity[filtered_variable_data_set$activity=="4"]<-"SITTING"
filtered_variable_data_set$activity[filtered_variable_data_set$activity=="5"]<-"STANDING"
filtered_variable_data_set$activity[filtered_variable_data_set$activity=="6"]<-"LAYING"
```

#Calculate the wide tidy data set
Now calculate the wide tidy data set by calculating the means for features.

```{r, echo=TRUE}

#get the unique/distinct combination of subject and activity
subject_activities <- unique(filtered_variable_data_set[c("subject","activity")])
cat(paste("Distinct combinations of subject and activity has ",ncol(subject_activities)," columns and ",nrow(subject_activities)," rows\n" ))

#iterate through the unique combination and calculate the means
tidy_data_set <-data.frame()
for(i in 1:nrow(subject_activities)) {
        row <- subject_activities[i,]
        this_subject=row[[1]]
        this_activity=row[[2]]
        
        #get only the entries for the current combination of subject and activity
        x<-filtered_variable_data_set[filtered_variable_data_set$subject==this_subject & filtered_variable_data_set$activity==this_activity,]
        drops <- c("subject","activity")
        x_num <- x[,!names(x) %in% drops]
        
        #calculate means
        averages <- colMeans(x_num)
        avg_v<- data.frame(as.list(averages))
        
        #create a new data frame with one single row containing the subject, activity and means 
        subject_and_activity <- data.frame(subject=this_subject, activity=this_activity)
        this_row <-cbind(subject_and_activity, avg_v)
        
        #rbind this row to the result data frame
        if(nrow(tidy_data_set)==0){
                tidy_data_set<-this_row
        }else{
                tidy_data_set<-rbind(tidy_data_set
                                     ,this_row)
        }
}
#name properly the columns
colnames(tidy_data_set)<-colnames(filtered_variable_data_set)
cat(paste("Tidy data set has  ",ncol(tidy_data_set)," columns and ",nrow(tidy_data_set)," rows\n" ))
```

##Final Step - write the file

```{r, echo=TRUE}
#write.table(tidy_data_set, file = "independent_tidy_data_set.txt", row.name=FALSE)
```
